### 2013-04-15 doc/final_report/evaluation.tex
I dun added mah fixes for amanda and maria's notes

Why write in first person in this chapter?

@@

Working with the Python implementation of SAPIENTA demonstrated the importance
of machine learning evaluation methods for testing classifiers rather than
making assumptions about `sensible looking' output. When the application was
run on a set of papers acquired for Partridge's corpus, it seemed to give
sensible CoreSC classifications for the papers. However, upon further analysis
and calculation of Python SAPIENTA's Classifier Accuracy, it was determined
that the system was only getting 44\% of CoreSC annotations correct, an
unacceptably low success rate. If I hadn't have run the evaluation functions on
SAPIENTA, it would have been easy to assume that it was correctly classifying
the papers, tainting the input papers and potentially disrupting the
investigation into CoreSC makeup and paper type.

\subsection{User Interface}
The Partridge user interface is also a great success, providing a highly usable
and intuitive representation of the relatively complicated processes that are
used to classify and organise the papers. The use of HTML and JavaScript
interactive elements instead of static pages and synchronous calls to the web
server was certainly a good idea, providing a modern system that keeps the user
informed at all times. The decision to use the jqBBQ library to add browser
history support further enhanced the user experience, allowing users of
Partridge to share links to specific search results over social media channels
and potentially bookmark significant results.

@@

Unfortunately, SAPIENTA is not scheduled for improvement work until the summer
and that means that Partridge's paper preprocessor is still not very
scalable. If further work to Partridge is carried out, I hope to improve the
accuracy of the Python SAPIENTA implementation to be comparable with the
current, experimental version of the system and run paper annotations locally
on the Partridge server.

@@

The relationship between a paper's  CoreSC makeup and its type has a lot of
potential for further research. During the development of Partridge, this
relationship was only tested using Research, Review and Case Study papers since
there were a large number such papers published in open access journals,
available for analysis. From the analysis in Section
\ref{sec:evaluaton_learners}, it is clear that a paper's CoreSC makeup can be
very effective in determining which of these three classes it belongs to. The
most discriminative features were specifically Background, Method and
Motivation. However, this relationship was only tested with the three paper
possible types listed above. Further work could be conducted to discern whether
the correlation between CoreSC makeup and paper type holds for other classes of
document and whether the most discriminative CoreSC labels remain the same for
a larger number of paper types.

>>>>

Why write in the first person in this chapter?

@@

Working with the Python implementation of SAPIENTA demonstrated the importance
of machine learning evaluation methods for testing classifiers, which are
discussed in depth in Chapter \ref{chapter:testing}, rather than making
assumptions about `sensible looking' output. Then the application was run on a
set of papers acquired for Partridge's corpus, it seemed to give sensible
CoreSC classifications for the papers. However, upon further analysis and
calculation of Python SAPIENTA's Classifier Accuracy, it was determined that
the system was only getting 44\% of CoreSC annotations correct, an unacceptably
low success rate. If I hadn't have run the evaluation functions on SAPIENTA, it
would have been easy to assume that it was correctly classifying the papers,
tainting the input papers and potentially disrupting the investigation into
CoreSC makeup and paper type.

\subsection{User Interface}

The Partridge user interface is also a great success, providing a highly usable
and intuitive representation of the relatively complicated processes that are
used to classify and organise the papers. The interface took a great deal of
tweaking and user feedback, as discussed in Section \ref{sec:user_testing}, in
order to make it as user friendly as possible. The use of HTML and JavaScript
interactive elements instead of static pages and synchronous calls to the web
server was certainly a good idea, providing a modern system that keeps the user
informed at all times.  The decision to use the jqBBQ library to add browser
history support further enhanced the user experience, allowing users of
Partridge to share links to specific search results over social media channels
and potentially bookmark significant results.

@@

Over the summer, I will be working with Dr Liakata to improve the Python
implementation of SAPIENTA. We aim to match or even surpass the performance of
the Perl SAPIENTA implementation and build a new web service using the Python
backend. It is possible that Partridge will use a local implementation of the
Python SAPIENTA implementation to process and annotate large volumes of new
papers as they are added to the corpus.

@@

The relationship between a paper's  CoreSC makeup and its type has a lot of
potential for further research. During the development of Partridge, this
relationship was only tested using Research, Review and Case Study papers since
there were a large number such papers published in open access journals,
available for analysis. From the analysis in Section
\ref{sec:evaluation_learners}, it is clear that a paper's CoreSC makeup can be
very effective in determining which of these three classes it belongs to. The
most discriminative features were specifically Background, Method and
Motivation. However, this relationship was only tested with the three paper
possible types listed above. Further work could be conducted to discern whether
the correlation between CoreSC makeup and paper type holds for other classes of
document and whether the most discriminative CoreSC labels remain the same for
a larger number of paper types.

### 2013-04-15 doc/final_report/implementation.tex
I dun added mah fixes for amanda and maria's notes

Implementing Partridge from a set of initial designs proved to be a
challenging but rewarding process. There were a great number of technical
challenges involved in building the system, a large proportion of the
implementation time was spent working on the paper preprocessor module.
However, some aspects of the web interface and backend were also quite
difficult to tackle.

Several libraries were used to accelerate the development and implementation of
Partridge and avoid `reinventing the wheel.' Whenever code libraries were used,
the role of the engineer was to be aware of the available solutions and deduce
which was aposite to the purpose intended.

@@

Partridge's Paper Preprocessor module was the most difficult part of the
project to implement. From acquiring papers for use in the project to teaching
a machine learning system to recognise different types of scientific
publication, a number of technical challenges were encountered.

\subsection{ Sourcing and Acquiring Scientific Papers}

In order to provide a useful service to researchers looking for papers, growing
a large corpus of scientific literature was made a high priority during the
project. This was especially important when training the Machine Learning
models since training data sets need to be statistically significant in order
to provide any meaningful results. Users are given the option, and actively
encouraged, to upload papers that they own or publish on another author's
behalf to the Partridge instance. However, to speed up this process, a number
of papers were acquired from various open access sources.

The most convenient and accessible source was the ART corpus that SAPIENTA was
trained with\cite{citeulike:11077287}. This corpus was already stored using the
CoreSC schema and had been pre-annotated. This meant that Partridge would not
need to do any conversion or pre-processing on the papers. The papers in this
corpus were all of a similar type and covered similar subject areas within the
domain of Biochemistry. There were also approximately 260 papers in the
collection. These papers were used to help establish a conversion process from
PDF to annotated XML. However, they only formed part of the final corpus.
Including a large number of papers from other sources provided a more
comprehensive collection of data for Partridge to learn from.

The `mega-journals' PLOSOne \url{http://www.plosone.org/} and arXiv
\url{http://www.arxiv.org/} were suggested as sources for more open access
articles that could be added to Partridge. Both of these sites were found to
contain large volumes of open access papers. Most of the articles stored on the
arXiv site were in PDF format which, as discussed below, are difficult to
convert and annotate. However, PLOSOne use the SciXML markup language for
papers published through their journal, which made converting and annotating
them a lot simpler for Patridge. PLOSOne also publish all of their papers under
the Creative Commons Share-Alike-By-Attribution license\cite{ccbyattr}. This
meant that as long as the author information was left intact, all of the papers
could be used for data mining purposes.

A third mega-journal used for downloading papers was the PubMed Centrala (PMC)
repository \url{http://www.ncbi.nlm.nih.gov/pmc/}. Not all of
the papers available for viewing at the PMC website are open access. However,
they do offer a listing for the subset of open access papers available on their
website. The PubMed Central format is also very similar to the SciXML format
and compatible with SAPIENTA. Therefore, a large number of PubMed papers were
also used for the initial training of Partridge.

\subsection{Format Conversion} Most scientific papers available on the internet
are formatted as PDF documents. However, Partridge uses and stores documents as
XML markup and uses the CoreSC schema by Soldatova and
Liakata\cite{liakata2008guidelines} for annotation. Therefore some spike work
was carried out to determine the feasibility of converting papers published as
PDF documents into XML documents. Townsend \emph{et al.} (2009) liken
converting PDF to XML to ``converting hamburgers into cows." They go on to
explain that PDF documents do not contain any semantic data and documents lose
much of their explicit structure when they are formatted in this way
\cite{Townsend2009}.  Therefore, to convert PDF documents into an NLP-friendly
format, some heuristics must be used to detect the document's
structure\cite{pdfminer}.

@@

The next stage was to split the text into sentences in preparation for
processing with SAPIENTA. With the assistance of the NLTK library, a sentence
splitting subroutine was implemented. This used a machine learning algorithm
that had been trained to recognise sentence boundaries to split the text. Each
sentence was then added to a CoreSC compatible XML document for processing by
SAPIENTA.

@@

Once all the data files and libraries were successfully in place, the Python
version of SAPIENTA was used to process some of the papers converted from PDF.
This appeared to have been successful. However, after applying some machine
learning evaluation techniques as discussed in Chapter \ref{chapter:testing},
it was clear that the Python version of SAPIENTA was not accurate enough to
provide CoreSC annotations, scoring 44\% accuracy on average. After discussion
with its author, it became apparent that some of the features provided in the
server-side version of SAPIENTA had not been implemented in the Python code
yet. Therefore, it was necessary to switch bak to remote annotation.

@@

Once a large number of papers had been acquired from ART, PlosOne and PubMed
Central, work could begin on implementing a paper type classifier able to
discriminate between `Case Study,' `Research' and `Review' papers. However,
for a machine learning classifier to discriminate between classes of data, it
is necessary to choose a set of features that differ as much as possible
between the classes. For example, to discriminate between cats and dogs,
``number of legs" is a poor feature since both species normally have four.
However, the noise that these animals make may be a good discriminative feature
since cats tend to mew and dogs bark.

@@

There are several well-established features used for classifying text within
the NLP community. However, since Partridge has access to CoreSC annotations
for each paper, an investigation was carried out into how useful CoreSC data
can be as a feature for paper type discrimination. Every sentence in a given
annotated paper has an associated CoreSC label. It is therefore trivial to
calculate the proportion of a paper made up of sentences with a specified
CoreSC type. Several papers were manually inspected and the proportions of each
CoreSC represented within were calculated. A pattern quickly emerged between
the types of paper and the proportions of each CoreSC within them.

@@

Before working with supervised learning systems, a preliminary experiment was
carried out to see whether clustering papers using the K-means algorithm could
successfully identify the paper types, thus removing the need to train and
maintain a supervised model. This was also an excellent opportunity to explore
the data and try and detect any inherent patterns within the papers.

K-means clustering uses the Euclidean distance between each data sample to
decide which cluster it is closest to and should belong to. K is the number of
clusters that are used and is decided ahead of execution. K-means clustering is
often run multiple times using a range of K values when the number of clusters
in the data is unknown. The performance of the clustering algorithm is often
evaluated using the K-means silhouette which is a measurement of how tightly
the data is clustered. A low silhouette value ($K \approx 0$) suggests loose
clustering and therefore little correlation, a high silhouette value ($K
\approx 1$) corresponds with very tight clustering and good correlation.

All of the papers were manually identified as either a `Case Study,' a
`Review Paper' or a `Research Paper.' They were then loaded into a script and
clustered using their CoreSC respective compositions. K-Means silhouettes were
used as a measure of how effective the clustering process was. This was
repeated for $ K \in [2..8] $ and the best silhouette measurement and value for
K retained. The algorithm consistently showed that K=3 was the best K value.
However, the silhouette value was consistently low. It was suggested that this
may have been because all CoreSC types were considered equally; the less
discriminative CoreSCs may have been offsetting the effect of the highly
discriminative ones.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.8\textwidth]{implementation/cluster.png}
\caption{Paper types rendered using Bac, Met and Mot CoreSCs}
\label{fig:clusters}
\end{figure}

@@

Clustering the papers did not seem to provide a reliable way to classify them
by type and it was decided that better results could be obtained through the
use of a supervised learning classifier instead. However, the three most
discriminative CoreSC types were chosen as features for the learner.

>>>>

Implementing Partridge from a set of initial designs proved to be a challenging
but rewarding process. Initially, a set of research papers to be used as for
Partridge were sourced from a number of open access journals and the Paper
Preprocessor system constructed. The backend infrastructure for the project was
then constructed and the web interface set up. After some initial suggestions
from early users, a web bookmarklet was developed to allow papers to be easily
imported from online journals such as PlOS ONE.

\section{ Sourcing and Acquiring Scientific Papers}

In order to provide a useful service to researchers looking for papers, growing
a large corpus of scientific literature was made a high priority during the
project. This was especially important when training the Machine Learning
models since training data sets need to be statistically significant in order
to provide any meaningful results. Users are given the option, and actively
encouraged, to upload papers that they own or publish on another author's
behalf to the Partridge instance. However, to speed up this process, a number
of papers were acquired from a variety of sources.

The most convenient and accessible source was the ART corpus that SAPIENTA was
trained with\cite{LIAKATA10.644}. This corpus was already stored using the
CoreSC schema and had been pre-annotated. This meant that Partridge would not
need to do any conversion or pre-processing on the papers. The papers in this
corpus were all of a similar type and covered similar subject areas within the
domain of Biochemistry. There were 265 papers in the collection. These papers
were used to help establish a conversion process from PDF to annotated XML.
However, they only formed part of the final corpus.  Including a large number
of papers from other sources provided a more comprehensive collection of data
for Partridge to learn from. A much larger number of papers had to be acquired
for training and testing of the machine learning systems in the paper
preprocessor.

Open Access (OA) journals are research journals that provide their papers free
to read online, often making revenue by charging authors to publish their
articles \cite{solomon2012study}. Many OA journals provide free read access to
their papers, but do not permit the user of articles for data mining purposes.
However, there are a number of large online journals that provide papers under
Creative Commons licenses, allowing data mining, provided that attribution of
the paper's author is provided in any derivative works. OA is a growing trend
and it is hoped that there will soon be enough permissively licensed, OA
literature generally available to allow Partridge to become a wide reaching and
useful research tool.

The `mega-journals' PLoSOne \url{http://www.plosone.org/} and arXiv
\url{http://www.arxiv.org/} were suggested as sources for more open access
articles that could be added to Partridge. Both of these sites were found to
contain large volumes of OA literature under a permissive license that allowed
data mining. Most of the articles stored on the arXiv site were in PDF format
which, as discussed below, are difficult to convert and annotate. However,
PLoSOne use the Pubmed Central markup schema, which is compatible with
SAPIENTA, for papers published through their journal, which made converting and
annotating them for inclusion into Partridge's corpus.  PLoSOne also publish
all of their papers under the Creative Commons Share-Alike-By-Attribution
license\cite{ccbyattr}. As mentioned above, this meant that as long as the
author information was left intact, all of the papers could be used for data
mining purposes.

A third mega-journal used for downloading papers was the PubMed Central (PMC)
repository \url{http://www.ncbi.nlm.nih.gov/pmc/}. Not all of
the papers available for viewing at the PMC website are open access. However,
they do offer a listing for the subset of OA papers available on their
website. The PubMed Central format is also very similar to the SciXML format
and compatible with SAPIENTA. Therefore, a large number of PubMed papers were
also used for the initial training of Partridge.

@@

With a number of papers available to use to form an initial corpus for the
system, work on the Paper Preprocessor could begin. Partridge's Paper
Preprocessor module was the most difficult part of the project to implement.
From acquiring papers for use in the project to teaching a machine learning
system to recognise different types of scientific publication, a number of
technical challenges were encountered.

\subsection{Format Conversion} Most scientific papers available on the internet
are formatted as PDF documents. However, Partridge uses and stores documents as
XML markup and uses the CoreSC
schema\cite{LIAKATA10.644}\cite{citeulike:10444769} for annotation.  Therefore
some spike work was carried out to determine the feasibility of converting
papers published as PDF documents into XML documents. Townsend \emph{et al.}
(2009) liken converting PDF to XML to ``converting hamburgers into cows." They
go on to explain that PDF documents do not contain any semantic data and
documents lose much of their explicit structure when they are formatted in this
way \cite{Townsend2009}.  Therefore, to convert PDF documents into an
NLP-friendly format, some heuristics must be used to detect the document's
structure\cite{pdfminer}.

@@

The next stage was to split the text into sentences in preparation for
processing with SAPIENTA, the classifier for CoreSC zones as discussed in
Section \ref{sec:nlpsapienta}. With the assistance of the NLTK library, a sentence
splitting subroutine was implemented. This used a machine learning algorithm
that had been trained to recognise sentence boundaries to split the text. Each
sentence was then added to a CoreSC compatible XML document for processing by
SAPIENTA.

@@

Once all the data files and libraries were successfully in place, the Python
version of SAPIENTA was used to process some of the papers converted from PDF.
This appeared to have been successful. However, after applying some machine
learning evaluation techniques as discussed in Chapter \ref{chapter:testing},
it was clear that the Python version of SAPIENTA was not accurate enough to
provide CoreSC annotations, scoring 44\% accuracy on average. After discussion
with its authors, it became apparent that some of the features provided in the
server-side version of SAPIENTA had not been implemented in the Python code
yet. It is intended that I will be employed to fix the Python SAPIENTA
implementation during the summer. However, until it is repaired, it was
necessary to switch back to remote annotation.

@@

Once a large number of papers had been acquired from ART, PlosOne and PubMed
Central, work could begin on implementing a paper type classifier able to
discriminate between `Case Study,' `Research' and `Review' papers. This would
allow researchers to filter papers into types, providing them with more search
precision and hopefully better results from using Partride.

For a machine learning classifier to discriminate between classes of data, it
is necessary to choose a set of features that differ as much as possible
between the classes.  For example, to discriminate between cats and dogs,
``number of legs" is a poor feature since both species normally have four.
However, the noise that these animals make may be a good discriminative feature
since cats tend to mew and dogs bark.

@@

There are several well-established features used for classifying text within
the NLP community. However, since Partridge has access to CoreSC annotations
for each paper, an investigation was carried out into how useful CoreSC data
can be as a feature for paper type discrimination. As discussed in Section
\ref{sec:nlpsapienta}, CoreSC information can be used to help users find
phrases within very specific contexts in the paper, vastly improving the
precision and accuracy of Partridge's search facility. Every sentence in a given
annotated paper has an associated CoreSC label.

It is therefore trivial to calculate the proportion of a paper made up of
sentences with a specified CoreSC type. Several papers were manually inspected
and the proportions of each CoreSC represented within were calculated. A
pattern quickly emerged between the types of paper and the proportions of each
CoreSC within them.

@@

Before working with supervised learning systems, a preliminary experiment was
carried out to see whether clustering papers using CoreSC tags as features for
the K-means algorithm could successfully identify the paper types, thus
removing the need to train and maintain a supervised model. This was also an
excellent opportunity to explore the data and try and detect any inherent
patterns within the papers.

K-means clustering uses the Euclidean distance between each data sample to
decide which cluster it is closest to and should belong to. K is the number of
clusters that are used and is decided ahead of execution. K-means clustering is
often run multiple times using a range of K values when the number of clusters
in the data is unknown.

The performance of the clustering algorithm is often
evaluated using the K-means silhouette which is a measurement of how tightly
the data is clustered\cite{Rousseeuw198753}. An $s(i)$ value is calculated for
each datum $i$ in the dataset. This is defined as:

\begin{displaymath}
s(i) = \frac{ b(i) - a(i) }{ max\{a(i), b(i)\} }
\end{displaymath}

For each $s(i)$, $a(i)$ denotes the dissimilarity of the data item to its
neighbours in the cluster it belongs to. It is defined using a distance
measure, in this case euclidean. Similarly, $b(i)$ denotes the lowest average dissimilarity of $i$ to
any cluster (i.e. neighbouring clusters too). This constrains $s(i)$ such that
$ -1 < s(i) < 1$. The average $s(i)$ for a cluster provides the cluster's
silhouette value and an indication of how appropriately data has been
clustered\cite{Rousseeuw198753}.

A low silhouette value ($K \approx
0$) suggests loose clustering and therefore little correlation, a high
silhouette value ($K \approx 1$) corresponds with very tight clustering and
good correlation.

All of the papers were manually identified as either a `Case Study,' a
`Review Paper' or a `Research Paper.' They were then loaded into a script and
clustered using their CoreSC respective compositions.

K-Means clustering relies upon the knowledge of how many clusters the data
should fit into before the algorithm starts. However silhouettes can be used to
evaluate a number of different $K$ values and determine which number of
clusters best fits the data. Since there are three classes, silhouettes were
used as a measure of how effective the clustering process was. This was
repeated for $ K \in [2..8] $ and the best silhouette measurement and value for
K retained. The algorithm consistently showed that K=3 was the best K value.
However, the silhouette value was consistently low. It was suggested that this
may have been because all CoreSC types were considered equally; the less
discriminative CoreSCs may have been offsetting the effect of the highly
discriminative ones.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.8\textwidth]{implementation/cluster.png}
\caption{Paper types rendered using Bac, Met and Mot CoreSCs; Research papers
are red, Review papers green and Case Studies blue.}
\label{fig:clusters}
\end{figure}

@@

Due to its unsupervised behaviour, Clustering can produce very unpredictable
results. Beyond specifying how many clusters to generate and the features to
use in discriminating between data items, the developer has no control over
behaviour of the k-means algorithm. In this case, there is no way to tell what
the three categories that have been generated correspond to. It was considered
that the clusters may correspond to the three main paper sources used: PloSONE,
Pubmed Central and the ART Corpus. However, the clusters did not match the
paper sources precisely either as can be seen in Figure
\ref{fig:cluster_sources}

\begin{figure}[!ht]
\centering
\includegraphics[width=0.8\textwidth]{implementation/cluster_sources.png}
\caption{Paper sources rendered using Bac, Met and Mot CoreSCs; PlosONE papers
are red, ART papers green and PubMed Central papers blue}
\label{fig:cluster_sources}
\end{figure}

Since clustering the papers did not seem to provide a reliable way to classify them
by type, it was decided that better results could be obtained through the
use of a supervised learning classifier instead. However, the three most
discriminative CoreSC types were chosen as features for the learner.
